The \dealii Library, Version 9.6

### 2024-09-23 9.6/paper.tex
Edit some more.

The major changes of this release are:
%
\begin{itemize}
  \item
    Substantial performance improvements to the matrix-free and multigrid
    infrastructure (Section~\ref{sec:mf}).
  \item
    Different additions to the non-matching infrastructure (see
    Section~\ref{sec:nonmatching}). In particular, the new class
    \texttt{FERemoteEvaluation} provides support for evaluating finite
    element shape functions and solutions on parts of the domain stored by
    other MPI processes.
  \item
    Much work has gone into writing wrappers for \trilinos' \tpetra{} stack
    of linear algebra classes. \tpetra{} is \trilinos' \kokkos-based
    replacement for the now-deprecated \epetra{} stack. See
    Section~\ref{sec:tpetra} for more on this.
  \item
    Tool classes \texttt{TaskResult} and \texttt{Lazy} that provide ways to
    compute values on a separate thread or only when first accessed
    (Section~\ref{sec:tools}).
  \item
    There are six new tutorial programs, on checkpointing simulations
    (\step{83}), integrating time-dependent solvers with external time
    stepping libraries (\step{86}, using \petsc's \ts{} library), advanced
    point evaluation techniques (\step{87}), non-matching grids (\step{89}),
    and trace-based methods for PDEs on embedded surfaces (\step{90}). See
    Section~\ref{subsec:steps} for more details.
  \item
    In Section~\ref{sec:license} we summarize the motivation and approach
    behind our relicensing effort to make \dealii available under the terms
    of the the Apache License 2.0 with LLVM Exception.
\end{itemize}
%

While all of these major changes are discussed in detail in
Section~\ref{sec:major}, there
are a number of other noteworthy changes in the current \dealii release,
which we briefly outline in the remainder of this section:
%
\begin{itemize}
  \item \dealii{} now requires and makes use of the C++17 language standard.
  \item We have continued to make progress in supporting simplex and
    mixed meshes -- mesh types that \dealii{} has traditionally not
    supported at all. Specifically, the current release uses
    better strategies for refinement of tetrahedra that result in
    better-shaped child cells. It also contains
    support for cubic finite elements on simplices.
  \item The \texttt{FE\_NedelecSZ} class that contains our
    implementation of the N\'ed\'elec element using the orientation
    scheme of \cite{Zag06} now supports the computation of hanging
    node constraints for locally refined meshes.
    \todo[inline]{Sebastian: Is there anything more to say here, maybe some reference? If
      you want, we can also list this point in the list above, and add
      a whole section 2.x for it.}
  \item The \texttt{AffineConstraints} class stores and processes
    constraints on degrees of freedom in \dealii{}. Such constraints
    can be of the form $x_3 = \frac 12 x_{14} + \frac 12 x_{15}$ as is
    common when using hanging node constraints (here, $x_3$ could
    represent the value of the solution at a hanging node, and $x_{14}$
    and $x_{15}$ are the values of the two adjacent degrees of freedom on
    the parent edge that contains the hanging node); or they can be of
    the form $x_{12}=42$ as is common when using Dirichlet boundary
    conditions.

    Since very early in the history of the library, the
    \texttt{AffineConstraints} class interface required building such
    constraints in multiple steps: First, one declared a degree of
    freedom as constrained; then one added the dependencies one after
    the other (e.g., by adding pairs $(\frac 12, 14)$ and $(\frac 12,
    15)$ in the hanging node example above); then one added
    inhomogeneities (by setting it to $42$ in the Dirichlet example
    above). This piecemeal approach is cumbersome and prevents the
    library from performing certain error checking steps because a
    constraint is not known to be completely built at any given point. The new
    \texttt{add\_constraint()} function now allows defining a
    constraint in one step.
  \item We refactored the systems for managing relative line and face
    orientations to significantly improve consistency across several library
    modules, including finite element classes, periodicity, and the
    \pfrst{} interface. \dealii{} has supported using unstructured
    three-dimensional meshes for a long time. Features like discontinuous
    Galerkin methods and higher-order elements require that lines and faces of
    adjacent elements are consistently oriented. For example, consider a line
    containing two degrees of freedom, which, in reference coordinates, are
    located at $x_1 = \frac 1 3$ and $x_2 = \frac 2 3$. If the line is read from
    left-to-right then we get points $(x_1, x_2)$ whereas if it is read from
    right-to-left we get $(x_2, x_1)$. \dealii{} guarantees that this order will
    be consistent across all cells sharing that line by storing a flag
    indicating whether (relative to the vertex enumeration on the present cell)
    that particular line is in the standard or reversed orientation. Faces in 3d
    are treated in a similar way. The only significant difference between lines
    and 3d faces is that, instead of two possible relative orientations,
    triangular faces have six and quadrilateral faces have eight.

    Historically, different places within the library implemented their own
    ad-hoc orientation logic, data encodings, and default values. We cleaned up
    this subsystem and instead of using three different booleans
    (\emph{orientation}, \emph{rotation}, and \emph{flip}), bitsets, or several
    other encodings, we use a single \texttt{unsigned char} which we named the
    \emph{combined orientation}. Since the representation in some library
    interfaces has changed, most of the old interfaces have been deprecated and
    a few (such as the low-level interface to periodic boundary conditions) had
    to be incompatibly altered or removed.
\item The \texttt{SolverGMRES} class now offers a third orthogonalization
  method, the classical Gram--Schmidt method with delayed orthogonalization
  \cite{Bielich2022}. Furthermore, the solver now specifies the maximal basis
  size of the Arnoldi basis, rather than the number of auxiliary
  vectors. Some changes have also been made to the GMRES and F-GMRES
  implementations, making them use the same underlying kernels as much as possible.
\item The new class \texttt{FE\_Hermite} implements a Hermite interpolation basis of
  maximum regularity. These bases are always of odd polynomial degree $p$ and have
  regularity $r=(p-1)/2$.
\end{itemize}
%
The changelog lists more than 180 features and bugfixes.

@@

This release of \dealii contains a number of large and significant changes,
which we will discuss in this section.
It of course also includes a
vast number of smaller changes and added functionality; the details of these
can be found
\href{https://dealii.org/developer/doxygen/deal.II/changes_between_9_5_2_and_9_6_0.html}
{in the file that lists all changes for this release}; see \cite{changes96}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Updates of multigrid and matrix-free algorithms}\label{sec:mf}

We made different updates of the multigrid and matrix-free infrastructure in \dealii. These include:
\begin{itemize}
\item Our own implementation of \texttt{std::simd}, called
  \texttt{VectorizedArray}, now also supports Arm Neon. Arm Neon is an
  architecture extension of the ARMv7, ARMv8 and ARMv9 architecture families,
  used, e.g., for the Arm Cortex-A and Arm Cortex-R series of processors or in
  the Apple 64-bit silicon series (A7--A18, M1--M4 as of 2024).
  With these instructions, 2 doubles or 4 floats can be processed in one
  go. Since the matrix-free infrastructure works directly with the \texttt{VectorizedArray}
  abstraction as data structure, it automatically benefits from this new implementation.

\item The application of the Piola transformation for values and gradients of
  $H$(div)-conforming Raviart--Thomas elements on non-Cartesian cells has
  been rewritten for better performance. Together with additional changes in
  the sum-factorization algorithms, which became more similar to the kernels
  described in~\cite{KronbichlerKormann2019}, the matrix-free operator
  evaluation is now three to five times faster in these cases.

@@

\item In addition, we performed substantial improvements to the global-coarsening
  multigrid infrastructure: \texttt{MGTransferMF} (previously:
  \texttt{MGTransferGlobalCoarsening}) and \texttt{MGTwoLevel\-Transfer}. They
  now allow to perform local smoothing, which is a key step towards unifying
  all transfer operators in \dealii. Finally, \texttt{MGTwoLevelTransfer} can now
  be set up with an existing \texttt{MatrixFree} object in the case
  of $p$-multigrid, allowing to reduce the setup costs and memory consumption significantly.
\end{itemize}

@@

In the non-matching infrastructure of \dealii, the following improvements were made:

\begin{itemize}
\item The performance of the non-nested multigrid infrastructure
(\texttt{MGTwoLevelTransferNon\-Nested}) has been improved significantly by avoiding
redundant copy operations. Furthermore, support for simplex-shaped cells and
multiple-component elements has been added.
\item Several minor performance improvements in \dealii's evaluator class on
  sets of unstructured points, \texttt{FEPointEvaluation}, have been
  made. This is particularly useful for evaluating several quantities on the
  same set of points, which is enabled by \texttt{NonMatching::MappingInfo}.
\item The new class \texttt{FERemoteEvaluation} has been added.
This is a class to access data
in a distributed matrix-free loop for non-matching discretizations.
Interfaces are named in analogy to \texttt{FEEvaluation}, in order to seamlessly switch the
local evaluator functionality at quadrature points of a computation. The key component is the
underlying MPI communication infrastructure, which is performed via \texttt{RemotePointEvaluation}.
Tutorial \step{89} has been added to present its usage
in the context of the application to acoustic conservation equations~\cite{heinz2023high}.
\item FECouplingValues

@@

Historically, \trilinos{} has implemented distributed linear algebra classes for vectors and (sparse) matrices in its
\epetra{} package that uses \MPI{} as its only source of
parallelism. Building on that, there are multiple interconnected \trilinos{} packages, e.g.\ for (non)-linear solvers 
and preconditioners, commonly referred to as the 
\epetra{} stack. For sake of brevity we will also refer to the whole stack as \epetra{} in the following.
As \epetra{} was the first stack, \dealii{} interfaces to \trilinos{} have traditionally 
been implemented to use it.

However, several years ago, \trilinos{} also introduced the
newer \tpetra{} (``templated'' Petra) package that provides additional shared memory
parallelism and GPU capabilities, referred to as \MPIx, by building on \kokkos{}. 
Since a few years, new features have only been implemented in \tpetra{} as 
it is slated to replace \epetra{}, and the latter is indeed
now deprecated with removal slated for 2025. 
As a consequence, 
we will eventually have to switch all of our \trilinos{} interfaces to \tpetra{}; 
the same is true for the need to switch to \tpetra{}-based
sub-packages replacing existing \trilinos{} sub-packages 
(for example, \texttt{Ifpack2} instead of \texttt{Ifpack}).

In the current release, we have put substantial work into this switch,
as outlined below. As this is not yet finished, it is important to note that 
all these features are optional and their individual availability
depends on the installed \trilinos{} packages. 
As an example, a \dealii{} build will pass with a \trilinos{} installation 
without \texttt{Ifpack2}, but its preconditioners will not be included.
In the following list, we omit the common namespace prefix
\texttt{LinearAlgebra::} on all mentioned
\texttt{TpetraWrappers} symbols for readability:
\begin{itemize}
  \item The \texttt{TpetraWrappers::Vector} class has been overhauled to also allow specifying
    a memory space, and the \texttt{TpetraWrappers::BlockVector} class has been added.
  \item \texttt{TpetraWrappers::SparseMatrix} and \texttt{TpetraWrappers::SparsityPattern} have been implemented mirroring
    the functionality of the \texttt{TrilinosWrappers::SparseMatrix} \\ and \texttt{TrilinosWrappers::SparsityPattern} classes.
  \item Various incomplete factorization and relaxation preconditioners (from \texttt{Ifpack2}) have been wrapped,
    mirroring the existing classes in \texttt{LinearAlgbera::TrilinosWrappers} as closely as possible.
  \item Two new preconditioner variants \texttt{TpetraWrappers::PreconditionL1Jacobi} and \\
    \texttt{TpetraWrappers::PreconditionL1GaussSeidel}, based on~\cite{BFKY2011}, have been added.
  \item \texttt{TpetraWrappers::SolverDirectKLU2} has been wrapped as a direct solver.
\end{itemize}
Some functionality in the \texttt{TrilinosWrappers} classes is still missing, 
most noticeably wrappers for the algebraic-multigrid preconditioner \texttt{MueLu}
and the iterative solvers from \texttt{Belos}.
However, the wrapped \texttt{Ifpack2} preconditioners can already be used with the iterative solvers of \dealii{}.

The solver and preconditioner classes mentioned above provide reasonable parameter subsets 
through \texttt{AdditionalData} objects, as in \texttt{TrilinosWrappers}. 
Additionally, there are new generic classes \texttt{TpetraWrappers::SolverDirect} 
and \texttt{TpetraWrappers::PreconditionIfpack} which expose the internal interface through
a \texttt{Teuchos::ParameterList} and thereby offering the full set of parameters for 
more experienced \trilinos{} users. This also allows the use of preconditioners or 
solvers not (yet) wrapped, e.g.\ \texttt{SuperLU\_dist} or \texttt{MUMPS}.

The design goal was to introduce as few changes as possible for the user to allow for an easy transition to the
\texttt{TpetraWrappers} classes.
However, based on the difference between \epetra{} and \tpetra{}, the following changes were necessary:
\begin{itemize}
  \item The constructor for the class \texttt{TrilinosWrappers::MPI::Vector} only required the \texttt{IndexSet locally\_owned},
    which describes the set of indices locally owned by the current rank.
    However, if one wants to create a vector that has read or write access to non-locally owned indices,  the constructor of the
    \texttt{TpetraWrappers::Vector} requires the \texttt{IndexSet locally\_owned}, the \texttt{IndexSet locally\_relevant}, and
    a boolean flag, whether to initialize the vector in the read-only or the write-only state.
  \item A vector created without providing the \texttt{IndexSet locally\_relevant} is purely local and cannot access non-local indices.
    Such a vector cannot be copied to a vector that can access non-local indices, as the \texttt{IndexSet locally\_relevant} must
    be provided at creating the vector object. 
  \item Some of solver and preconditioner parameters are not available in \tpetra{},
    such that the \texttt{AdditionalData} objects are not identical. 
    Since there are many preconditioners we will not list each individual change but instead refer 
    to our Doxygen documentation. 
  \item The incomplete Cholesky (IC) factorization preconditioner is not available 
   in \tpetra{}.
\end{itemize}

@@

One other important change is that
all the \texttt{TpetraWrappers} classes take \texttt{Number} and \texttt{MemorySpace} template arguments,
which control the underlying number type and memory space.
By default, the memory is allocated on the CPU, mirroring the behavior of the \texttt{TrilinosWrappers}.
The support for different number types also enables the usage of automatic differentiation-based tools in
\trilinos{}, such as \texttt{NOX}.
Additionally, the memory space type determines whether \tpetra{} can utilize GPUs if \kokkos{} is configured
with a GPU backend.
To make the automatic differentiation tools of \trilinos{} accessible through the \tpetra{} interface without
requiring users to have an in-depth understanding of \trilinos{}, additional wrappers for the corresponding
\trilinos{} packages will be added in future releases.
Furthermore, to fully leverage the GPU backend and unlock its full potential for users will require further work.

@@

  (C++ provides functionality via \texttt{std::async} with
  launch policy \texttt{std::launch::deferred} that can achieve
  similar outcomes. But this functionality is more difficult to use than
  \texttt{Lazy<T>} because, among other reasons, the code generating the
  object has to be specified at the place of construction of the
  object holding the result, rather than at the place of use; and
  because the holder object -- \texttt{std::future} -- can only be
  asked once for its computed value.)

@@

Many of the \dealii tutorial programs were revised in a variety of ways
as part of this release. In addition, there are a number of new tutorial
programs:
\begin{itemize}
  \item
    \step{83}
    demonstrates how one can implement
    checkpoint/restart functionality in \dealii-based programs,
    using the \boost{} serialization functionality as a
    foundation. \step{83} was written by Pasquale Africa, Wolfgang
    Bangerth, and Bruno Blais using \step{19} as its basis.
    \todo{Also add affiliations like for the other tutorials?}
  \item
    \step{86}
    is a program that solves the heat equation using \petsc's \ts{} (time
    stepping) framework for the solution of ordinary differential
    equations. Written by Wolfgang Bangerth (Colorado State
    University), Luca Heltai (University of Pisa), and Stefano Zampini
    (King Abdullah University of Science and Technology), it
    illustrates how PDE solvers for time-dependent problems can be
    integrated with existing ODE solver packages to use advanced ODE
    solver concepts (such as higher-order time integration methods and
    adaptive time step control), all without sacrificing the things
    that have traditionally led code authors toward writing their own
    time stepping routines (such as wanting to change the mesh every
    once in a while, or having to deal with boundary conditions).
  \item
    \step{87}~\cite{schreter2023step87} was contributed by Magdalena Schreter-Fleischhacker
    (Technical University of Munich) and Peter Munch
    (University of Augsburg/Uppsala University). It
    presents the advanced point-evaluation functionalities of \dealii,
    which are useful for evaluating finite element solutions at
    arbitrary points on meshes that can be distributed among processes. The presented mini-examples are motivated by the application to two-phase-flow simulations and demonstrate, for example, the evaluation of solution quantities at a surface mesh embedded in a background mesh, as is the case in front-tracking.
  \item
    \step{89} was contributed by Johannes Heinz (TU Wien),
    Maximilian Bergbauer (Technical University of Munich),
    Marco Feder (SISSA), and Peter Munch (University of Augsburg/Uppsala University).
    It shows a way how to apply non-matching and/or Chimera methods
    within matrix-free loops in \dealii.
  \item
    \step{90} was contributed by Vladimir Yushutin and Timo Heister (Clemson University).
    It implements the trace finite element method (TraceFEM). TraceFEM solves PDEs
    posed on a, possibly evolving, $(dim-1)$-dimensional surface $\Gamma$ employing
    a fixed uniform background mesh of a $dim$-dimensional domain in which
    the surface is embedded. Such surface PDEs arise in problems involving
    material films with complex properties and in other situations in which
    a non-trivial condition is imposed on either a stationary or a moving interface.
    The program considers a steady, complex, non-trivial surface and the prototypical
    Laplace-Beltrami equation which is a counterpart of
    the Poisson problem on flat domains.
\end{itemize}

>>>>

The major changes of this release are:
%
\begin{itemize}
  \item
    Substantial performance improvements to the matrix-free and multigrid
    infrastructure (Section~\ref{sec:mf}).
  \item
    Additions to the non-matching infrastructure (see
    Section~\ref{sec:nonmatching}). In particular, the new class
    \texttt{FERemoteEvaluation} provides support for evaluating finite
    element shape functions and solutions on parts of the domain stored by
    other MPI processes.
  \item
    Much work has gone into writing wrappers for \trilinos' \tpetra{} stack
    of linear algebra classes. \tpetra{} is \trilinos' \kokkos-based
    replacement for the now-deprecated \epetra{} stack. See
    Section~\ref{sec:tpetra} for more on this.
  \item
    Tool classes \texttt{TaskResult} and \texttt{Lazy} that provide ways to
    compute values on a separate thread or only when first accessed
    (Section~\ref{sec:tools}).
  \item
    There are six new tutorial programs, on checkpointing simulations
    (\step{83}), integrating time-dependent solvers with external time
    stepping libraries (\step{86}, using \petsc's \ts{} library), advanced
    point evaluation techniques (\step{87}), non-matching grids (\step{89}),
    and trace-based methods for PDEs on embedded surfaces (\step{90}). See
    Section~\ref{subsec:steps} for more details.
  \item
    In Section~\ref{sec:license} we summarize the motivation and approach
    behind our relicensing effort to make \dealii available under the terms
    of the the Apache License 2.0 with LLVM Exception.
\end{itemize}
%

While all of these major changes are discussed in detail in
Section~\ref{sec:major}, there
are a number of other noteworthy changes in the current \dealii release,
which we briefly outline in the remainder of this section:
%
\begin{itemize}
  \item \dealii{} now requires and makes use of the C++17 language standard.
  \item We have continued to make progress in supporting simplex and
    mixed meshes -- mesh types that \dealii{} has traditionally not
    supported at all. Specifically, the current release uses
    better strategies for refinement of tetrahedra that result in
    better-shaped child cells. It also contains
    support for cubic finite elements on simplices.
  \item The \texttt{FE\_NedelecSZ} class that contains our
    implementation of the N\'ed\'elec element using the orientation
    scheme of \cite{Zag06} now supports the computation of hanging
    node constraints for locally refined meshes.
    \todo[inline]{Sebastian: Is there anything more to say here, maybe some reference? If
      you want, we can also list this point in the list above, and add
      a whole section 2.x for it.}
  \item The \texttt{AffineConstraints} class stores and processes
    constraints on degrees of freedom in \dealii{}. Such constraints
    can be of the (homogeneous) form $x_3 = \frac 12 x_{14} + \frac 12 x_{15}$ as is
    common when using hanging node constraints (here, $x_3$ could
    represent the value of the solution at a hanging node, and $x_{14}$
    and $x_{15}$ are the values of the two adjacent degrees of freedom on
    the parent edge that contains the hanging node); or they can be of
    the form $x_{12}=42$ as is common when using Dirichlet boundary
    the (inhomogeneous) form $x_{12}=42$ as is common when using Dirichlet boundary
    conditions.

    Since very early in the history of the library, the
    \texttt{AffineConstraints} class interface required building such
    constraints in multiple steps: First, one declared a degree of
    freedom as constrained; then one added the dependencies one after
    the other (e.g., by adding pairs $(\frac 12, 14)$ and $(\frac 12,
    15)$ in the hanging node example above); then one added
    inhomogeneities (by setting it to $42$ in the Dirichlet example
    above). This piecemeal approach is cumbersome and prevents the
    library from performing certain error checking steps because a
    constraint is not known to be completely built at any given point. The new
    \texttt{add\_constraint()} function now allows defining a
    constraint in one step.
  \item We refactored the systems for managing relative line and face
    orientations to significantly improve consistency across several library
    modules, including finite element classes, periodicity, and the
    \pfrst{} interface. \dealii{} has supported using unstructured
    three-dimensional meshes for a long time. Features like discontinuous
    Galerkin methods and higher-order elements require that lines and faces of
    adjacent elements are consistently oriented. For example, consider a line
    containing two degrees of freedom, which, in reference coordinates, are
    located at $x_1 = \frac 1 3$ and $x_2 = \frac 2 3$. If the line is read from
    left-to-right then we get points $(x_1, x_2)$ whereas if it is read from
    right-to-left we get $(x_2, x_1)$. \dealii{} guarantees that this order will
    be consistent across all cells sharing that line by storing a flag
    indicating whether (relative to the vertex enumeration on the present cell)
    that particular line is in the standard or reversed orientation. Faces in 3d
    are treated in a similar way. The only significant difference between lines
    and 3d faces is that, instead of two possible relative orientations,
    triangular faces have six and quadrilateral faces have eight
    possible orientation.

    Historically, different places within the library implemented their own
    ad-hoc orientation logic, data encodings, and default values. We cleaned up
    this subsystem and instead of using three different booleans
    (\emph{orientation}, \emph{rotation}, and \emph{flip}), bitsets, or several
    other encodings, we now use a single \texttt{unsigned char} which we named the
    \emph{combined orientation}. Since the representation in some library
    interfaces has changed, most of the old interfaces have been deprecated and
    a few (such as the low-level interface to periodic boundary conditions) had
    to be incompatibly altered or removed.
\item The \texttt{SolverGMRES} class now offers a third orthogonalization
  method, the classical Gram--Schmidt method with delayed orthogonalization
  \cite{Bielich2022}. Furthermore, the solver now specifies the maximal basis
  size of the Arnoldi basis, rather than the number of auxiliary
  vectors. Some changes have also been made to the GMRES and F-GMRES
  implementations, making them use the same underlying kernels as much as possible.
\item The new class \texttt{FE\_Hermite} implements a Hermite interpolation basis of
  maximum regularity. These bases are always of odd polynomial degree $p$ and have
  regularity $r=(p-1)/2$.
\end{itemize}
%
The
\href{https://dealii.org/developer/doxygen/deal.II/changes_between_9_5_2_and_9_6_0.html}{changelog}
-- listing more than 180 features and bugfixes --
contains a complete record of all changes; see \cite{changes96}.

@@

This release of \dealii contains a number of large and significant changes,
which we will discuss in this section.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Updates to the multigrid and matrix-free algorithms}\label{sec:mf}

We updated a number of significant parts of the multigrid and matrix-free
infrastructure in \dealii. These changes include:
\begin{itemize}
\item Our own implementation of the \texttt{std::experimental::simd}
  class proposed for inclusion into the C++ standard, called
  \texttt{VectorizedArray}, now also supports Arm Neon. Arm Neon is an
  architecture extension of the ARMv7, ARMv8 and ARMv9 architecture families,
  used, e.g., for the Arm Cortex-A and Arm Cortex-R series of processors or in
  the Apple 64-bit silicon series (A7--A18, M1--M4 as of 2024).
  With these instructions, 2 doubles or 4 floats can be processed in one
  go. Since the matrix-free infrastructure works directly with the \texttt{VectorizedArray}
  abstraction as data structure, it automatically benefits from this new implementation.

\item The application of the Piola transformation for values and gradients of
  $H$(div)-conforming Raviart--Thomas elements on non-Cartesian cells has
  been rewritten for better performance. Together with additional changes in
  the sum-factorization algorithms, which became more similar to the kernels
  described in~\cite{KronbichlerKormann2019}, the matrix-free operator
  evaluation is now 3--5 times faster in these cases.

@@

\item In addition, we performed substantial improvements to the global-coarsening
  multigrid infrastructure: \texttt{MGTransferMF} (previously:
  \texttt{MGTransferGlobalCoarsening}) and \texttt{MGTwoLevel\-Transfer}. They
  now allow performing local smoothing, which is a key step towards unifying
  all transfer operators in \dealii. Finally, \texttt{MGTwoLevelTransfer} can now
  be set up with an existing \texttt{MatrixFree} object in the case
  of $p$-multigrid, reducing the setup costs and memory consumption significantly.
\end{itemize}

@@

In the non-matching infrastructure of \dealii, we made the following improvements:

\begin{itemize}
\item The performance of the non-nested multigrid infrastructure
(\texttt{MGTwoLevelTransferNon\-Nested}) has been improved significantly by avoiding
redundant copy operations. Furthermore, support for simplex-shaped cells and
multiple-component elements has been added.
\item Several minor performance improvements in \dealii's evaluator class on
  sets of unstructured points, \texttt{FEPointEvaluation}, have been
  made. This is particularly useful for evaluating several quantities on the
  same set of points, which is enabled by \texttt{NonMatching::MappingInfo}.
\item We added the new class \texttt{FERemoteEvaluation}.
This is a class to access data
in a distributed matrix-free loop for non-matching discretizations.
Interfaces are named in analogy to \texttt{FEEvaluation}, in order to seamlessly switch the
local evaluator functionality at quadrature points of a computation. The key component is the
underlying MPI communication infrastructure, which is performed via \texttt{RemotePointEvaluation}.
Tutorial \step{89} has been added to present its usage
in the context of the application to acoustic conservation equations~\cite{heinz2023high}.
\item FECouplingValues

@@

Historically, \trilinos{} has implemented distributed linear algebra classes for vectors and (sparse) matrices in its
\epetra{} package that uses \MPI{} as its only source of
parallelism. Building on that, there are multiple interconnected \trilinos{} packages, e.g.\ for (non-)linear solvers 
and preconditioners, commonly referred to as the 
``\epetra{} stack''. For sake of brevity we will also refer to the whole stack as \epetra{} in the following.
\dealii{}'s interfaces to \trilinos{} have traditionally 
been implemented to use it.

However, several years ago, \trilinos{} also introduced the
newer \tpetra{} (``templated'' Petra) package that provides additional shared memory
parallelism and GPU capabilities, referred to as \MPIx, by building on \kokkos{}. 
In recent years, new features have only been implemented in \tpetra{} as 
it is slated to replace \epetra{}, and the latter is indeed
now deprecated with removal slated for 2025. 
As a consequence, 
we will eventually have to switch all of our \trilinos{} interfaces to \tpetra{}; 
the same is true for the need to switch to \tpetra{}-based
sub-packages (the ``Tpetra stack'') that are intended to replace existing \trilinos{} sub-packages 
(for example, \texttt{Ifpack2} instead of \texttt{Ifpack}).

In the current release, we have put substantial work into this switch,
as outlined below. As this is not yet finished, it is important to note that 
all these features are optional and their individual availability
depends on the installed \trilinos{} packages. 
As an example, a \dealii{} build will pass with a \trilinos{} installation 
without \texttt{Ifpack2}, but its preconditioners will not be included.
In the following list, we omit the common namespace prefix
\texttt{LinearAlgebra::} on all mentioned
\texttt{TpetraWrappers} symbols for readability:
\begin{itemize}
  \item The \texttt{TpetraWrappers::Vector} class has been overhauled to also allow specifying
    a memory space, and the \texttt{TpetraWrappers::BlockVector} class has been added.
  \item \texttt{TpetraWrappers::SparseMatrix} and \texttt{TpetraWrappers::SparsityPattern} have been implemented mirroring
    the functionality of the \texttt{TrilinosWrappers::SparseMatrix} \\ and \texttt{TrilinosWrappers::SparsityPattern} classes.
  \item Various incomplete factorization and relaxation preconditioners (from \texttt{Ifpack2}) have been wrapped,
    mirroring the existing classes in \texttt{LinearAlgbera::TrilinosWrappers} as closely as possible.
  \item Two new preconditioner variants \texttt{TpetraWrappers::PreconditionL1Jacobi} and \\
    \texttt{TpetraWrappers::PreconditionL1GaussSeidel}, based on~\cite{BFKY2011}, have been added.
  \item \texttt{TpetraWrappers::SolverDirectKLU2} is now available as a direct solver.
\end{itemize}
Some functionality in these wrapper classes is still missing, 
most noticeably wrappers for the algebraic-multigrid preconditioner \texttt{MueLu}
and the iterative solvers from \texttt{Belos}.
However, the wrapped \texttt{Ifpack2} preconditioners can already be used with the iterative solvers of \dealii{}.

The solver and preconditioner classes mentioned above provide reasonable parameter subsets 
through \texttt{AdditionalData} objects, just as the existing
\texttt{Epetra} wrappers. 
Additionally, there are new generic classes \texttt{TpetraWrappers::SolverDirect} 
and \texttt{TpetraWrappers::PreconditionIfpack} which expose the internal interface through
a \texttt{Teuchos::ParameterList} and thereby offering the full set of parameters for 
more experienced \trilinos{} users. This also allows the use of preconditioners or 
solvers not (yet) wrapped, e.g.\ \texttt{SuperLU\_dist} or \texttt{MUMPS}.

The design goal was to introduce as few changes as possible for the user to allow for an easy transition to the
\texttt{TpetraWrappers} classes.
However, based on the difference between \epetra{} and \tpetra{}, the following changes were necessary:
\begin{itemize}
  \item The constructor for the class \texttt{TrilinosWrappers::MPI::Vector} only required the \texttt{IndexSet locally\_owned},
    which describes the set of indices locally owned by the current rank.
    However, if one wants to create a vector that has read or write access to non-locally owned indices,  the constructor of the
    \texttt{TpetraWrappers::Vector} requires the \texttt{IndexSet locally\_owned}, the \texttt{IndexSet locally\_relevant}, and
    a boolean flag, whether to initialize the vector in the read-only or the write-only state.
  \item A vector created without providing the \texttt{IndexSet locally\_relevant} is purely local and cannot access non-local indices.
    Such a vector cannot be copied to a vector that can access non-local indices, as the \texttt{IndexSet locally\_relevant} must
    be provided at creating the vector object. 
  \item Some of parameters of existing solver and preconditioner are
    not available in the \tpetra{} wrappers,
    such that the corresponding \texttt{AdditionalData} objects are not identical. 
    Since there are many preconditioners we will not list each individual change but instead refer 
    to our Doxygen documentation. 
  \item The incomplete Cholesky (IC) factorization preconditioner is not available 
   in \tpetra{}.
\end{itemize}

@@

Finally,
all \texttt{TpetraWrappers} classes take \texttt{Number} and \texttt{MemorySpace} template arguments,
in order to control the underlying scalar type and memory space.
By default, memory is allocated on the CPU, mirroring the behavior of
the \texttt{TrilinosWrappers}, but choosing a different memory space
also allows \tpetra{} to utilize GPUs if \kokkos{} is configured
with a GPU backend.
Furthermore, to fully leverage the GPU backend and unlock its full potential for users will require further work.
The support for different number types in vector and matrix classes also enables the usage of automatic differentiation-based tools in
\trilinos{}, such as \texttt{NOX}.
To make the automatic differentiation tools of \trilinos{} accessible through the \tpetra{} interface without
requiring users to have an in-depth understanding of \trilinos{}, additional wrappers for the corresponding
\trilinos{} packages will be added in future releases.

@@

  (C++ provides functionality via \texttt{std::async} with
  launch policy \texttt{std::launch::deferred} that can achieve
  similar outcomes. But this functionality is more difficult to use than
  \texttt{Lazy<T>} because, among other reasons, the code generating the
  object has to be specified at the place of construction of the
  object holding the result, rather than at the place of use; and
  because the holder object -- \texttt{std::future} -- can only be
  asked \textit{once} for its computed value.)

@@

Many of the \dealii tutorial programs were revised in a variety of ways
as part of this release: Around 190 of the more than 1900 (non-merge)
commits that went into this release touched the tutorial.
% data generated using these commands:
% - tutorial: git log --since 2023/07/07 --until 2024/08/11 --no-merges examples | grep commit | wc -l
% - total:    git log --since 2023/07/07 --until 2024/08/11 --no-merges | grep commit | wc -l
In addition, there are a number of new tutorial
programs:
\begin{itemize}
  \item
    \step{83}
    demonstrates how one can implement
    checkpoint/restart functionality in \dealii-based programs,
    using the \boost{} serialization functionality as a
    foundation. \step{83} was written by Pasquale Africa, Wolfgang
    Bangerth, and Bruno Blais using \step{19} as its basis.
  \item
    \step{86}
    is a program that solves the heat equation using \petsc's \ts{} (time
    stepping) framework for the solution of ordinary differential
    equations. Written by Wolfgang Bangerth, Luca Heltai, and Stefano Zampini
    (King Abdullah University of Science and Technology), it
    illustrates how PDE solvers for time-dependent problems can be
    integrated with existing ODE solver packages to use advanced ODE
    solver concepts (such as higher-order time integration methods and
    adaptive time step control), all without sacrificing the things
    that have traditionally led code authors toward writing their own
    time stepping routines (like wanting to change the mesh every
    once in a while, or having to deal with boundary conditions).
  \item
    \step{87}~\cite{schreter2023step87} was contributed by Magdalena Schreter-Fleischhacker
    and Peter Munch. It
    presents the advanced point-evaluation functionalities of \dealii,
    which are useful for evaluating finite element solutions at
    arbitrary points on meshes that can be distributed among
    processes. The presented mini-examples are motivated by the
    application to two-phase flow simulations and demonstrate, for
    example, the evaluation of solution quantities at a surface mesh
    embedded in a background mesh, as needed in the case in front-tracking.
  \item
    \step{89} was contributed by Johannes Heinz (TU Wien),
    Maximilian Bergbauer (Technical University of Munich),
    Marco Feder (SISSA), and Peter Munch.
    It shows how to apply non-matching and/or Chimera methods
    within matrix-free loops in \dealii.
  \item
    \step{90} was contributed by Vladimir Yushutin (Clemson University) and Timo Heister.
    It implements the trace finite element method (TraceFEM). TraceFEM solves PDEs
    posed on a, possibly evolving, $(dim-1)$-dimensional surface $\Gamma$ employing
    a fixed uniform background mesh of a $dim$-dimensional domain in which
    the surface is embedded. Such surface PDEs arise in problems involving
    material films with complex properties and in other situations in which
    a non-trivial condition is imposed on either a stationary or a moving interface.
    The program considers a steady, complex, non-trivial surface and the prototypical
    Laplace-Beltrami equation which is a counterpart of
    the Poisson problem on flat domains.
\end{itemize}
